Classless Java: Tuning Java Interfaces

### 2015-12-07 paper/sections/CaseStudy.tex
revised bib and lots of small changes
compressed to 25 pages (toggled out author notes) again.

\paragraph{Initial System}
In the formulation of the EP there is an initial system that models
arithmetic expressions with only literals and addition, and an initial 
operation \texttt{eval} for expression evaluation.
As shown in Figure~\ref{fig:ep}, \texttt{Exp} is the common
super-interface with an evaluation operation \texttt{eval()}
inside. Sub-interfaces \texttt{Lit} and \texttt{Add} extend interface
\texttt{Exp} with default implementations for the \texttt{eval} operation. The
number field \texttt{x} of a literal is represented as a getter method
\texttt{x()} and expression fields (\texttt{e1} and \texttt{e2}) of an addition
as getter methods \texttt{e1()} and \texttt{e2()}.

@@

%\marco{text under need to be improved}
\paragraph{Adding a New Operation} The difficulty of the EP in OO  
languages arises from adding new operations. For example, adding a pretty printing 
operation would typically change all existing code. However, a solution 
%to the EP forbids this and it also forbids using casts. In other words, it should be possible 
should add the pretty printing operation in a type-safe and modular way. This
turns out to be easily achieved with the assistance of \mixin.  The code in
Figure~\ref{fig:ep} (on the right) shows how to add the new operation \texttt{print}.
Interface \texttt{ExpP} extends \texttt{Exp} with the extra method
\texttt{print()}. Interfaces \texttt{LitP} and \texttt{AddP} are defined with
default implementations of \texttt{print()}, extending base interfaces
\texttt{Lit} and \texttt{Add}, respectively. Importantly, note that in
\texttt{AddP}, the types of ``\emph{fields}'' (i.e. the getter methods)
\texttt{e1} and \texttt{e2} are refined. If the types were not refined then
the \texttt{print()} method in \texttt{AddP} would fail to type-check.

@@

\lstinputlisting[linerange=37-37]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_QUERY2
\marco{put on two lines}

>>>>

\paragraph{Initial System}
In the formulation of the EP there is an initial system that models
arithmetic expressions with only literals and addition, and an initial 
operation \texttt{eval} for expression evaluation.
As shown in Figure~\ref{fig:ep}, \texttt{Exp} is the common
super-interface with operation \texttt{eval()}
inside. Sub-interfaces \texttt{Lit} and \texttt{Add} extend interface
\texttt{Exp} with default implementations for the \texttt{eval} operation. The
number field \texttt{x} of a literal is represented as a getter method
\texttt{x()} and expression fields (\texttt{e1} and \texttt{e2}) of an addition
as getter methods \texttt{e1()} and \texttt{e2()}.

@@

%\marco{text under need to be improved}
\paragraph{Adding a New Operation} The difficulty of the EP in OO  
languages arises from adding new operations. For example, adding a pretty printing 
operation would typically change all existing code. However, a solution 
%to the EP forbids this and it also forbids using casts. In other words, it should be possible 
should add operations in a type-safe and modular way. This
turns out to be easily achieved with the assistance of \mixin.  The code in
Figure~\ref{fig:ep} (on the right) shows how to add the new operation \texttt{print}.
Interface \texttt{ExpP} extends \texttt{Exp} with the extra method
\texttt{print()}. Interfaces \texttt{LitP} and \texttt{AddP} are defined with
default implementations of \texttt{print()}, extending base interfaces
\texttt{Lit} and \texttt{Add}, respectively. Importantly, note that in
\texttt{AddP}, the types of ``\emph{fields}'' (i.e. the getter methods)
\texttt{e1} and \texttt{e2} are refined. If the types were not refined then
the \texttt{print()} method in \texttt{AddP} would fail to type-check.

@@

\lstinputlisting[linerange=37-38]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_QUERY2

### 2015-12-07 paper/sections/Conclusion.tex
compressed to 25 pages (toggled out author notes) again.

Before Java 8, concrete methods and static methods were not allowed
to appear in interfaces.  Java 8 allows static interface methods and
introduces \emph{default methods}, which enables implementations
inside interfaces. This had an important positive consequence that
was probably overlooked: the concept of class
(in Java) is now (almost) redundant and unneeded.  We define a subset of Java,
called Classless Java, where programs and (reusable) libraries can be
easily defined and used.  To avoid syntactic boilerplate
caused by Java not being originally designed to work without classes,
we introduce a new annotation, \mixin, which provides default implementations
for various methods (e.g. getters, setters, with-methods) and a
mechanism to instantiate objects. The \mixin annotation helps programmers
to write less cumbersome code while coding in Classless Java. Indeed
we think the obtained gain is so high that Classless Java with the \mixin
annotation can be less cumbersome than full Java.
Interestingly, without classes there is also no subclassing. This scratches an old
  itching point in the long struggle of subtyping versus subclassing:
  According to some authors, from a software engineering perspective,
  interfaces are just a kind of classes. Others consider more
  opportune to consider interfaces as pure types. We do not know how to conciliate
  those two viewpoints and Classless Java design.
  Classless Java does not have classes purely in the Java sense.  
 Classless Java encourages coding in a more flexible way by either
 keeping a higher abstraction level (interfaces are a more abstract
 concept than classes), or relying on concrete object initialization
 (the \Q@new I(){...}@ construct).

>>>>

Before Java 8, concrete methods and static methods were not allowed
to appear in interfaces.  Java 8 allows static interface methods and
introduces \emph{default methods}, which enables implementations
inside interfaces. This had an important positive consequence that
was probably overlooked: the concept of class
(in Java) is now (almost) redundant and unneeded.  We define a subset of Java,
called Classless Java, where programs and (reusable) libraries can be
easily defined and used.  To avoid syntactic boilerplate
caused by Java not being originally designed to work without classes,
we introduce a new annotation, \mixin, which provides default implementations
for various methods (e.g. getters, setters, with-methods) and a
mechanism to instantiate objects. The \mixin annotation helps programmers
to write less cumbersome code while coding in Classless Java. Indeed
we think the obtained gain is so high that Classless Java with the \mixin
annotation can be less cumbersome than full Java.
Interestingly, without classes there is also no subclassing. This scratches an old
  itching point in the long struggle of subtyping versus subclassing:
  according to some authors, from a software engineering perspective,
  interfaces are just a kind of classes. Others consider more
  opportune to consider interfaces as pure types. We do not know how to conciliate
  those two viewpoints and Classless Java design.
  Classless Java does not have classes purely in the Java sense.  
 Classless Java encourages coding in a more flexible way by either
 keeping a higher abstraction level (interfaces are a more abstract
 concept than classes), or relying on concrete object initialization
 (the \Q@new I(){...}@ construct).

### 2015-12-07 paper/sections/Implementation.tex
compressed to 25 pages (toggled out author notes) again.

The annotation we created is \mixin. In Eclipse, with an interface annotated by
\mixin, the automatic annotation processing is performed transparently and the information of
the interface from compilation is captured in the ``Outline'' window. This includes
all the methods inside the interface as well as the generated ones.  The custom
transformation is easy and convenient to use.  For example this means that the
IDE functionality for content assist and autocomplete will work for the newly generated
methods. The biggest reasons to use Lombok rather than using a conventional Java
 annotation processor are:
\begin{itemize}
%this seamd like a commercial spot
%\item Lombok is byte-code based instead of source-code based, which makes client
%  code concise and easy to maintain. Such code generation is performed at
%  compile time to modify bytecode.
\item Lombok modifies the generation process of the class files, by directly modifying the AST.
Neither the source code is modified nor new Java files are generated.
\item Moreover, and probably more importantly, Lombok is capable of generating
  code \emph{inside} a class/interface. This is the ability that conventional
  Java annotation processors do not provide.
\end{itemize}

>>>>

The annotation we created is \mixin. In Eclipse, with an interface annotated by
\mixin, the automatic annotation processing is performed transparently and the information of
the interface from compilation is captured in the ``Outline'' window. This includes
all the methods inside the interface as well as the generated ones.  The custom
transformation is easy and convenient to use.  For example this means that the
IDE functionality for content assist and autocomplete will work for the newly generated
methods. The biggest reasons to use Lombok rather than using a conventional Java
 annotation processor are:
\begin{itemize}
%this seamd like a commercial spot
%\item Lombok is byte-code based instead of source-code based, which makes client
%  code concise and easy to maintain. Such code generation is performed at
%  compile time to modify bytecode.
\item Lombok modifies the generation process of the class files, by directly modifying the AST.
Neither the source code is modified nor new Java files are generated.
\item Moreover, and probably more importantly, Lombok is capable of generating
  code \emph{inside} a class/interface, which conventional
  Java annotation processors do not support.
\end{itemize}

### 2015-12-07 paper/sections/Overview.tex
compressed to 25 pages (toggled out author notes) again.

\noindent will be accepted and behave as expected. If the programmer wishes to suppress
this behaviour and keep the signature as it was, it is sufficient to redefine
the \Q@with-@ methods in the new interface repeating the old signature.  Again,
the philosophy is that if the programmer provides something directly, \mixin
does not touch it.  The cast in \Q@with(Point2D)@ is trivially safe since it is
guarded by an \Q@instanceof@ test. The idea is that if the parameter is a
subtype of the current exact type, then we can just return the parameter, as
something that is just ``more'' than \Q@this@.

>>>>

\noindent will be accepted. If the programmer wishes to suppress this behaviour
and keep the signature as it was, it is sufficient to redefine the \Q@with-@
methods in the new interface repeating the old signature.  Again, the philosophy
is that if the programmer provides something directly, \mixin does not touch it.
The cast in \Q@with(Point2D)@ is trivially safe because of the \Q@instanceof@
test. The idea is that if the parameter is a subtype of the current exact type,
then we can just return the parameter, as something that is just ``more'' than
\Q@this@.

### 2015-12-07 paper/sections/RelatedWork.tex
compressed to 25 pages (toggled out author notes) again.

\paragraph{Mixins.}
Mixins are a more restricted model than the C++ approach. Mixins allow to name
components that can be applied to various classes as reusable functionality
units. However, % they suffer from linearisation: the order of mixin application
% is relevant in often subtle and undesired ways. This hinders their usability
% and the ability of resolving conflicts:
the linearisation (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in some cases and
restricts the flexibility of mixin composition. To fight this limitation, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raises the complexity, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits are in fact more like linearised mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearised mixin composition. Other languages, such as
Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@
\end{comment}
\noindent Java interfaces and default methods do not use
linearisation: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

@@

There are also proposals for extending Java with traits. For
example, FeatherTrait Java (FTJ)~\cite{Liquori08ftj} extends
FJ~\cite{Igarashi01FJ} with statically-typed traits, adding trait-based
inheritance in Java.  Except for few, mostly syntactic details, their work can
be emulated with Java 8 interfaces. There are also extensions to the original
trait model, which have operations that default methods and interfaces cannot
model, such as method renaming~\cite{reppy2006foundation}, which breaks
structural subtyping.

@@

Other languages offer even stronger support for safe code manipulation:
Template Haskell~\cite{sheard2002template}, F\# (type providers)~\footnote{http://research.microsoft.com/fsharp/} and MetaFjig (Active Libraries)~\cite{servetto2010metafjig}
all allow to execute code at compile time. They generate code/classes that is transparently integrated in the program that is being generated/processed/compiled. In particular, MetaFjig offers a property called \emph{meta-level-soundness}. In short this property ensures by construction that library code (even if wrong or unreasonable) never generates ill-typed code. This is roughly equivalent to
\textit{Self coherence}, that we have to manually prove.
Since MetaFjig is not working on annotated classes, there is no direct
equivalence on the overall theorem of the shown safety.\bruno{Don't understand
the end of the sentence!}

>>>>

\paragraph{Mixins.}
Mixins are a more restricted model than the C++ approach. Mixins allow to name
components that can be applied to various classes as reusable functionality
units. However, % they suffer from linearisation: the order of mixin application
% is relevant in often subtle and undesired ways. This hinders their usability
% and the ability of resolving conflicts:
the linearisation (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in some cases and
restricts the flexibility of mixin composition. To fight this limitation, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raises the complexity, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits are in fact more like linearised mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearised mixin composition. % Other languages, such as
% Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@
\end{comment}
\noindent Java interfaces and default methods do not use
linearisation: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

@@

There are also proposals for extending Java with traits. For
example, FeatherTrait Java (FTJ)~\cite{Liquori08ftj} extends
FJ~\cite{Igarashi01FJ} with statically-typed traits, adding trait-based
inheritance in Java.  Except for few, mostly syntactic details, their work can
be emulated with Java 8 interfaces. There are also extensions to the original
trait model, with operations (e.g. renaming~\cite{reppy2006foundation}, which breaks
structural subtyping) that default methods and interfaces cannot
model.

@@

Other languages offer even stronger support for safe code manipulation:
Template Haskell~\cite{sheard2002template}, F\# (type providers)~\footnote{http://research.microsoft.com/fsharp/} and MetaFjig (Active Libraries)~\cite{servetto2010metafjig}
all allow to execute code at compile time. They generate code that is
transparently integrated in the program that is being
generated/processed/compiled. In particular, MetaFjig offers a property called
\emph{meta-level-soundness}, ensuring by construction that library code (even if wrong or unreasonable) never generates ill-typed code. This is roughly equivalent to
\textit{Self coherence}, that we have to manually prove.
Since MetaFjig is not working on annotated classes, there is no direct
equivalence on the overall theorem of the shown safety.\bruno{Don't understand
the end of the sentence!}
